---
title: "Data Visualization in R - Workshop"
author: "Ha Dinh"
date: '2018-04-30'
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This workshop acquires some materials from the [Data Wrangling workshop](https://github.com/davidklaing/data_science_workshops) led by David Laing. 

## How to get set up for this workshop

### Getting this document on your computer:

1. Go to the GitHub repository here: [https://github.com/hadinh1306/workshop-materials](https://github.com/hadinh1306/workshop-materials)
2. Click the green button on the right that says "Clone or download".
3. Click "Download ZIP". (If you're proficient with git, feel free to clone the repository.)
4. Create a folder on your computer to store your work, and store your ZIP file there.
5. Double-click your ZIP file to unzip it and get all the code.

### Getting R and Rstudio

1. Download and install R from here: [http://cran.stat.sfu.ca/](http://cran.stat.sfu.ca/).
2. Download and install RStudio Desktop (Open Source Edition) from here: [https://www.rstudio.com/products/rstudio/#Desktop](https://www.rstudio.com/products/rstudio/#Desktop).

### Getting ready to play!

1. In RStudio, open `data_viz_workshop.Rmd`, a file in `YOUR_FOLDER/workshop_materials/data_visualization`. (That's this file!)
2. In the code snippet below, remove the hashtags from both lines, and click the green "play" button on the right to install `dplyr`, `ggplot2`, and `titanic`, the three packages you'll need for the workshop.

```{r}
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("gapminder")
```

### Having installation problems?

Option 1: Ask for help! We have volunteers who can help troubleshoot.

Option 2a: Find a partner and follow along together — most of the exercises can be done collaboratively. Your installation problem is almost certainly solvable — we just might not have time today.

Option 2b: Use [this datacamp light](https://cdn.datacamp.com/dcl/standalone-example.html) page to write and execute code in your browser. (A couple caveats: doesn't come with the titanic data, and might end up distracting you from the lesson. If this is your first time programming, I recommend option 2a.)

## Let's start!
### What is *Data Visualization*? 

Data visualization is a presentation of data in graphical, pictoral format. It helps you find trends in data, or tell an appealing story from the data you have.

We will work with tabular data - arranged in rows and column, like we did in the [Data Wrangling workshop](https://github.com/davidklaing/data_science_workshops). We are going to learn some fundamental components of `ggplot2` to create a visualization:
- Set up a plot with `ggplot`
- Choose which variables to plot using `mapping = aes()`
- Choose which type of plot using `geom_`
- Add title and subtitle using `labels`

### An inspiration!
Hans Rosling's [They Joyful of Stat](https://youtu.be/jbkSRLYSojo?t=33s) is one of my inspirations for data visualization. He turned boring numbers into an engaging story using data visualization! Today we might not re-create his visualization completely, but you will have a taste of how using R `ggplot2` package.


